Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-02 _tex/paper2.tex
Built site for gh-pages

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines Causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ experiments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

@@

A motivating example that will appear in the rest of the document
clarifies these concepts. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Following current CJ practices,
researchers rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that Thurstone's Case V, and by extension the BTL model, suffers from
several statistical and measurement limitations. These limitations
hinder the model's ability to identify various estimands relevant to CJ
inquiries, including the one described in the example. Identification is
crucial because it is a necessary condition for ensuring consistent
estimators. \emph{Consistency} refers to the property of an estimator
whose estimates converge to the ``true'' value of the estimand as the
data size approaches infinity \citep{Everitt_et_al_2010}. Without
identification, consistency cannot be achieved, even with ``infinite''
and error-free data. Thus, deriving meaningful insights from finite data
becomes impossible \citep{Schuessler_et_al_2023}.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to Causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as follows:
\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
\vdots \\
i \\
\vdots \\
h \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
i & a \\
\vdots & \vdots \\
h & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

Here, each element of \(I\) represents a unique individual \(i\) or
\(h\), where \(n_{I}\) denotes the total number of individuals.
Similarly, each element of \(J\) corresponds to a unique judge \(j\),
with \(n_{J}\) indicating the total number of judges. Moreover, each row
of \(IA\) represents a unique pairing of individuals \(i, h\) with
stimuli \(a, b\). As a result, the matrix \(IA\) contains
\(n_{I} \cdot n_{A}\) rows and \(2\) columns, where \(n_{A}\) specifies
the number of stimuli available per individual. Likewise, each row of
\(JK\) associates a judge \(j\) with a (repeated) judgment index \(k\).
Consequently, the matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\)
columns, where \(n_{K}\) indicates the number of repeated judgments each
judge makes. \begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

Additionally, we construct the matrix \(R\) to map each row of the
\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
matrix has \(n\) rows and \(6\) columns, where
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
coefficient, which quantifies the total number of unique comparisons
possible between every pair of stimuli generated by each pair of
individuals in the population. Thus, we define the matrix as in Equation
(\ref{eq-mat12}).

It is easier to visualize the structure of the previously defined
vectors and matrices by considering an example. Assuming \(n_{I} = 5\),
\(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), the vectors and
matrices described in Equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the form as in Equation (\ref{eq-mat13}).
\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

@@

We can illustrate the structure of these vectors more clearly with an
example. Suppose researchers exclude the second student, the second text
from each student, and the third judge from the setup shown in Equation
(\ref{eq-mat13}). Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
\(n_{K} = 3\), the resulting vectors would have the following structure:
\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
1 \\
1 \\
1 \\
1 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in Equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in Equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.
\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

@@

Thus, to address these limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines Causal and Bayesian inference methods.
The approach begins by formulating a conceptual model represented by a
Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ experimental design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes in the CJ system.

@@

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in CJ experimental designs, one that better aligns with the
demands of contemporary CJ assessment contexts \citep{Kelly_et_al_2022},
offering new insights into existing research and opening promising
avenues for future inquiry.

Five research avenues, both current and new, deserve particular
attention. The following sections outline these avenues and explain how
our model provides a rigorous research framework to investigate them.
Notably, although it may seem like a natural next step to address these
questions through formal derivations, following the tradition of
classical statistical models like linear regression, we contend that the
complexity of the CJ system makes this approach impractical from the
outset. Instead, simulation-based methods, such as power analysis, offer
a more feasible and flexible alternative, enabling researchers to
investigate these questions without depending on cumbersome mathematical
proofs. Nonetheless, developing formal mathematical proofs remains an
essential goal for future work.

@@

Despite its promise, this study presents several challenges that
researchers must navigate --challenges common to many quantitative
studies. These challenges fall into two main categories. The first
concerns the background knowledge researchers need to apply the proposed
framework effectively. The second involves the increased attention
researchers must now devote to their specific CJ research context.

In terms of background knowledge, researchers face two challenges:
mastering Causal inference and developing proficiency in Bayesian
inference methods. One example illustrating the need to master Causal
inference is our assumption that predictor variables are ``relevant'' to
the research context--interpreting relevance as the inclusion of these
variables in a ``sufficient adjustment set''
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this entails or its implications for model
specification, estimation, and inference. As a result, researchers must
engage deeply with these complex ideas--and with others, such as how
SCMs encode functional and probabilistic information--to apply the
framework rigorously. To assist this effort, we provide key references
throughout the study to guide motivated readers toward a deeper
understanding of these foundational concepts\footnote{refer to footnote
  1, 2, and the detailed online document referenced in the Declarations
  section}.

Bayesian inference presents a second challenge. Researchers must acquire
sufficient expertise to translate context-specific conceptual models
into bespoke statistical models that meaningfully address their research
questions. Although such translations are not unique to Bayesian
approaches, we argue that Bayesian methods offer a more accessible
route, as they reduce the need for specialized knowledge in optimization
theory and techniques--often required by frequentist methods--thereby
lowering the technical barriers for many applied researchers.

In terms of the attention researchers must now devote to their specific
CJ research context, researchers face two challenges: ensuring the
conceptual model is a \emph{faithful} representation for the CJ system,
and verifying whether this model can accurately estimate the intended
estimand from real data.

@@

Beyond their graphical capabilities, SCMs and DAGs can encode the
probabilistic information embedded within a causal structure. They
achieve this encoding by relying on three fundamental assumptions: the
local Markov, the minimality, the causal edges assumption. The
\emph{local Markov assumption} encodes probabilistic independencies
between variables by declaring that nodes in a graph are independent of
all its non-descendants, given its parents \citep[pp.~20]{Neal_2020}.
Meanwhile, the \emph{minimality assumption} encodes probabilistic
dependencies among variables by stating that every pair of adjacent
nodes exhibits a dependency \citep[pp.~21]{Neal_2020}. Finally, the
\emph{causal edges assumption} encodes causal relationships between
variables by declaring that each parent node acts as a direct cause of
its children \citep[pp.~22]{Neal_2020}. Figure~\ref{fig-ACflow}
illustrates how these assumptions influence the statistical and causal
interpretations of graphs.

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{images/png/ACflow.png}

\caption{\label{fig-ACflow}The flow of association and causation in
graphs. Extracted and slightly modified from \citet[pp.~31]{Neal_2020}}

>>>>

To address these issues, this study extends Thurstone's general form
through a systematic and integrated approach that combines causal and
Bayesian inference methods. In addition to potentially enhancing
measurement reliability and validity, and improving statistical accuracy
in hypothesis testing, this approach offers two key advantages. First,
it clarifies the interactions among all actors and processes involved in
CJ experiments. Second, it shifts the current comparative data analysis
paradigm from passively accepting the BTL model assumptions to actively
testing whether those assumptions fit the data under analysis.

@@

A motivating example that will appear throughout this document clarifies
these concepts. In this example, researchers aim to answer the question:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this, a
researcher designs a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges then compare pairs of students' written texts (stimuli) to
produce a dichotomous outcome reflecting the relative quality of each
text (trait). Based on this setup, researchers can reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''.

Following current CJ practices, researchers rely on estimates from the
BTL model--or its transformations--to approximate this estimand.
However, as discussed in Section~\ref{sec-theory-issues}, Thurstone's
Case V and the BTL model exhibit several statistical and measurement
limitations. These limitations hinder the model's ability to identify
various estimands relevant to CJ inquiries, including the one described
in the example.

@@

Thus, this section addresses the issues identified in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
combines the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to improving statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

@@

Although specifying a data structure is not mandatory when using SCMs
and DAGs, defining one in this case can improve clarity and facilitate
the description of the system. Thus, to re-express the scalar form of
the CJ system shown in Figure~\ref{fig-cj03} into an equivalent
vectorized form, we first define the vectors \(I\) and \(J\), along with
the matrices \(IA\) and \(JK\), as in Equation (\ref{eq-mat11}). Here,
each element of \(I\) represents a unique individual \(i\) or \(h\),
where \(n_{I}\) denotes the total number of individuals. Similarly, each
element of \(J\) corresponds to a unique judge \(j\), with \(n_{J}\)
indicating the total number of judges. Moreover, each row of \(IA\)
represents a unique pairing of individuals \(i, h\) with stimuli
\(a, b\). As a result, the matrix \(IA\) contains \(n_{I} \cdot n_{A}\)
rows and \(2\) columns, where \(n_{A}\) specifies the number of stimuli
available per individual. Likewise, each row of \(JK\) associates a
judge \(j\) with a (repeated) judgment index \(k\). Consequently, the
matrix \(JK\) has \(n_{J} \cdot n_{K}\) rows and \(2\) columns, where
\(n_{K}\) indicates the number of repeated judgments each judge makes.
\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
\vdots \\
i \\
\vdots \\
h \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
i & a \\
\vdots & \vdots \\
h & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A}
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

Additionally, we construct the matrix \(R\) to map each row of the
\(IA\) matrix with a corresponding row from the \(JK\) matrix. This
matrix has \(n\) rows and \(6\) columns, where
\(n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}\). Here, the
term \({n_{I} \cdot n_{A} \choose 2}\) represents the binomial
coefficient, which quantifies the total number of unique comparisons
possible between every pair of stimuli generated by each pair of
individuals in the population. Thus, we define the matrix as in Equation
(\ref{eq-mat12}). \begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
i & a & h & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & 1 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{A}-1 & n_{I} & n_{A} & n_{J} & n_{K}
\end{bmatrix}
}\end{equation}

It is easier to visualize the structure of the previously defined
vectors and matrices by considering an example. Assuming \(n_{I} = 5\),
\(n_{A} = 2\), \(n_{J} = 3\), and \(n_{K} = 3\), the vectors and
matrices described in Equations (\ref{eq-mat11}) and (\ref{eq-mat12})
take the form as in Equation (\ref{eq-mat13}).

\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 2 & 1 & 1 \\
1 & 1 & 5 & 2 & 1 & 2 \\
1 & 1 & 5 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
4 & 2 & 5 & 2 & 3 & 1 \\
4 & 2 & 5 & 2 & 3 & 2 \\
4 & 2 & 5 & 2 & 3 & 3 \\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 
\end{bmatrix}
}\end{equation}

@@

We can illustrate the structure of these vectors more clearly with an
example. Suppose researchers exclude the second student, the second text
from each student, and the third judge from the setup shown in Equation
(\ref{eq-mat13}). Given \(n_{I} = 5\), \(n_{A} = 2\), \(n_{J} = 3\), and
\(n_{K} = 3\), the resulting vectors would have the following structure:

\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
1 \\
1 \\
1 \\
1 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

@@

Finally, we define the \emph{sample mechanism} \(S\) in Equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Each element \(s_{(i,a,h,b,j,k)}\) is a binary value
indicating the presence or absence of data rows in the matrix \(R\)
resulting from the sample mechanism, as in Equation (\ref{eq-mat26}).
Thus, the vector contains \(n\) elements, matching the number of rows in
\(R\), and the sum of its elements represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

@@

\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1,1,1,2,1,1)} \\
\vdots \\
s_{(1,1,1,2,1,n_{K})} \\
\vdots \\
s_{(i,a,h,b,j,k)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)} \\
\vdots \\
s_{(n_{I},n_{A}-1,n_{I},n_{A},n_{J},1)}
\end{bmatrix}
}\end{equation}

@@

Thus, to address these limitations of Thurstone's Case V and the BTL
model, this study extends Thurstone's general form using a systematic,
integrated approach that combines causal and Bayesian inference methods.
The approach begins by formulating a conceptual model represented by a
Structural Causal Model (SCM) and a Directed Acyclic Graph (DAG). This
model integrates Thurstone's core theoretical principles, such as the
discriminal processes of stimuli, alongside key CJ experimental design
features, including judges' bias, sampling procedures, and comparison
mechanisms, thereby disentangling the causal processes in the CJ system.

@@

Beyond these potential benefits, the approach offers two additional
advantages. First, it clarifies the roles and interactions of all actors
and processes involved in CJ experiments. Second, it shifts the analytic
paradigm from passively accepting the assumptions of the BTL model to
actively testing their fit with observed data. Together, these
advantages establish a principled framework for evaluating best
practices in CJ experimental designs, one that better aligns with the
demands of contemporary CJ assessment contexts \citep{Kelly_et_al_2022},
offering new insights into existing research and opening promising
avenues for future inquiry. Five current and new research avenues
deserve particular attention due to their direct impact on the
reliability and validity of trait estimates and the accuracy of
statistical inferences. The following sections outline these avenues and
explain how our model provides a rigorous research framework to
investigate them.

@@

Despite its promise, this study presents several challenges that
researchers must navigate--challenges common to many quantitative
studies. They fall into two main categories. The first concerns the
foundational knowledge researchers need to apply the proposed framework
effectively. The second involves the increased attention researchers
must now devote to their specific research context.

\subsubsection{Foundational knowledge}\label{sec-discussion_challenges1}

Researchers must develop foundational knowledge in two critical areas:
causal and Bayesian inference methods. One example that illustrates the
importance of mastering causal inference is our common assumption that
predictor variables are ``relevant'' to the research
context--interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this assumption entails or its implications for
model specification, estimation, and inference. As a result, to apply
the framework rigorously, researchers must engage deeply with these and
other complex ideas, such as how SCMs encode functional and
probabilistic information. To assist this effort, we have provided key
references throughout the study to guide motivated readers toward a
deeper understanding of these foundational concepts\footnote{refer to
  footnote 1, 2, and the detailed online document referenced in the
  Declarations section}.

Additionally, researchers must develop the expertise to translate
context-specific conceptual models into bespoke statistical models using
Bayesian inference methods. While Bayesian approaches introduce specific
technical demands--such as familiarity with probabilistic programming
languages (PPLs) and concepts like probability distributions and
convergence--we argue they offer a more accessible path for many applied
researchers. This is because Bayesian methods significantly reduce the
need for specialized knowledge in areas like optimization theory and
techniques, which are often necessary in frequentist approaches.

\subsubsection{Attention to the research
context}\label{sec-discussion_challenges2}

When applying our framework to a specific CJ research context,
researchers face two additional challenges. First, they must verify that
the conceptual model provides a \emph{faithful} \footnote{Avoid
  confusing this term with the \emph{faithfulness assumption} described
  in the causal inference literature. For further details, see footnote
  1.} representation of the CJ system under study. Second, they must
evaluate whether the model's statistical translation can accurately
estimate the intended estimands (i.e., parameters) from empirical data.
Notably, addressing these challenges requires careful attention because
causal inference relies on expert knowledge and assumptions about the
variables' causal structure that are often untestable at the outset
\citep{Hernan_et_al_2025}. Consequently, the framework does not deliver
automatic answers; instead, it encourages researchers to formulate
precise questions and explicitly state their assumptions to build a
generalizable understanding of the phenomenon under study, in this case,
the CJ system \citep{Rohrer_et_al_2022, Deffner_et_al_2022}. This
theory-driven foundation then supports the empirical testing of causal
assumptions \citep{Deffner_et_al_2022}.

Thus, regarding the \emph{faithful} representation of the CJ system, we
advise researchers to treat our conceptual and statistical models as
starting points rather than universal solutions for all CJ designs or
datasets. This consideration means that researchers must face the
challenge of adapting these models to their specific research contexts,
paying close attention to causal assumptions, experimental design, and
model specification. This careful adaptation is essential because a
model's ability to produce accurate estimates and valid inferences
depends heavily on how well the data and inferential goals align with
its underlying assumptions \citep{Kohler_et_al_2019}. Nevertheless, as
\citet{Kohler_et_al_2019} points out, drawing conclusions about the real
world from observed data always requires assumptions, regardless of
whether the data are observational or experimental. In this context, our
approach provides a transparent framework for researchers to clearly
articulate these assumptions and design CJ experiments that support
reliable and valid trait measurement.

Finally, regarding the estimation capabilities of the statistical model,
researchers must tackle the challenge of performing identification
analysis. As outlined in Section~\ref{sec-theoretical},
\emph{identification analysis} determines whether a statistical model
can accurately compute a given estimand (e.g., a parameter) based solely
on its (causal) assumptions, independent of random variability
\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
necessary condition for consistency. \emph{Consistency} is the property
of an estimator (or model) whose estimates converge to the ``true''
value of an estimand as data size approaches infinity
\citep{Everitt_et_al_2010}. Logically, if researchers fail to identify a
model, it cannot achieve consistency--even with infinite, error-free
data. As a result, deriving meaningful insights from finite data becomes
impossible \citep{Schuessler_et_al_2023}.

Although performing identification analysis through formal derivations
may appear to be a natural next step, we argue that the complexity of
the CJ system renders such an approach impractical at the outset.
Instead, simulation-based methods, such as power analysis, provide a
more practical and flexible alternative, enabling researchers to examine
the consistency of estimates without relying on cumbersome mathematical
proofs. Nonetheless, the development of formal identification proofs
remains a significant goal for future CJ research.

@@

Beyond their graphical capabilities, researchers can translate SCMs and
DAGs into statistical models because these conceptual models encode
functional and probabilistic information, which they can then replace
with suitable functions and probabilistic assumptions
\citep{Pearl_et_al_2016}. Notably, this encoding relies on two
fundamental assumptions: the causal Markov and the faithfulness
assumption. The \emph{causal Markov assumption} links the graph
structure to the joint probability distribution of variables by
asserting that any variable in a graph is independent of all its
non-descendants conditional on its parents
\citetext{\citealp[pp.~20]{Neal_2020}; \citealp[pp.~80]{Hernan_et_al_2025}}.
The \emph{faithfulness assumption}, in turn, states that the observed
statistical (in)dependencies in the data reflect the underlying causal
structure among variables--except in rare degenerate cases
\citetext{\citealp[pp.~100]{Neal_2020}; \citealp[pp.~81]{Hernan_et_al_2025}}.
